---
author: "Jocelyne Horanituze"
date: "Sep 10, 2026"
output:
  github_document:
    keep_html: true
    pandoc_args: ["--self-contained", "--mathjax"]
always_allow_html: true
---
**Jocelyne Horanituze**

Sep 10, 2026

This section works through the nycflights13 dataset end to end: airport reference data, departure delay summaries, air time distributions, top/summer-only/high-altitude destinations, cancellation rates mapped by destination, and how departure delay, arrival delay, wind speed, and time of day relate to one another.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(dplyr)
library(ggplot2)
library(nycflights13)
library(ggridges)
library(tidyr)
library(ggExtra)

nyc_Airports <- unique(flights$origin)
nyc_flights <- filter(flights, origin %in% nyc_Airports)
```

## 1. Airport Reference

```{r, echo=FALSE, message=FALSE}
d <- data.frame( filter(airports, faa %in% nyc_Airports) %>%
                select(name, faa))
kbl <- knitr::kable(d, col.names = c("Airport Name", "Airport Code"),
                    caption= "New York City Airport Names" ,format = "html" )
kableExtra::kable_styling(kbl, full_width = FALSE)
```

## 2. Departure Delay Summary

```{r, message=FALSE, echo=FALSE}
d <- data.frame(nyc_flights %>%
               group_by(origin) %>%
               summarise(
                 avg_dep_delay = mean(dep_delay, na.rm = TRUE),
                 med_dep_delay = median(dep_delay, na.rm = TRUE))
               )

kbl <- knitr::kable(d, col.names = c("Airport Code", "Average dep delay", "Median dep delay"),
                    caption= "Average and Median Departure Delays" ,format = "html" )
kableExtra::kable_styling(kbl, full_width = FALSE)
```

- EWR has the highest average departure delay, followed by JFK and LGA, with delays of 15.1 minutes, 12.1 minutes, and 10.3 minutes respectively.
- The median delay tells a different story: LGA has the highest median delay, followed by JFK and then EWR, with median delays of -3 minutes, -1 minute, and -1 minute respectively.
- Even though EWR has the highest average departure delay, it's more likely to delay by fewer minutes compared to LGA and JFK.

## 3. Air Time Distributions

```{r, message=FALSE, warning=FALSE, echo=FALSE}
ggplot(nyc_flights, aes(x = air_time, color = origin),
       alpha = 0.5) +
  geom_density() +
  labs(title = "Single plot using color and alpha")+
  theme_get()

ggplot(nyc_flights, aes(x = air_time, fill = origin)) +
    geom_density(bw = 1, alpha = 0.5) +
    facet_wrap(~origin, nrow = 1) +
  labs(title = "Faceted plot")

ggplot(nyc_flights, aes(x = air_time, y = origin)) +
    geom_density_ridges(aes(height = after_stat(density)),
                        stat = "density", bw = 1) +
  labs(title = "Ridgeline plot")
```

- **Overall**
    - The single plot and ridgeline plot help compare the distributions of flights from all airports.
    - The distributions of flights from EWR and JFK are fairly similar, with their tails extending to around 400 minutes, whereas LGA doesn't extend past 200 minutes. Overall there is a mix of destinations from each airport.
    - The faceted plot, with one density per facet, shows a clearer distribution for each airport's flights, but makes comparisons between airports harder.
    - For the single plot, if fill is used instead of color, the overlapping distributions become hard to tell apart.

## 4. Top Destinations

```{r, echo=FALSE, message=FALSE}
# top four destinations with the most flights from New York City
top_dest <- count(flights, dest, sort = TRUE) %>%
                    slice_max(n = 4, order_by = n)

top_flights <- flights %>%
                filter(dest %in% top_dest$dest)

# proportion of flights from each of the three New York City airports to top destinations
propp <- top_flights %>%
  count(dest, origin) %>%
  group_by(dest) %>%
  mutate(prop = n / sum(n)) %>%
  ungroup()

ggplot(propp) +
  geom_col(aes(x = origin, y = prop, fill = origin)) +
  facet_wrap(~dest, ncol = 2) +
  labs(title = "Proportions of Flights from New York City Airports to Top Destinations",
       x = "Origin Airport",
       y = "Proportion of Flights") +
  theme_get()
```

## 5. Summer-Only Destinations

```{r, echo=FALSE, message=FALSE}
summer_flights <- flights %>%
          filter(month %in% c(6,7,8))
non_summer_flights <- flights %>%
          filter(!month %in% c(6,7,8))

# destinations only served during summer months
summer_flights_dest <- anti_join(summer_flights, non_summer_flights, "dest") %>%
                        distinct(dest)

summer_airports <- semi_join(airports, summer_flights_dest, by = c("faa" = "dest")) %>%
                   select(faa, name)

kbl <- knitr::kable(summer_airports, col.names = c("Airport code", "Airport name"),format = "html" )
kableExtra::kable_styling(kbl, full_width = FALSE)
```

## 6. High Altitude Destinations

```{r, echo=FALSE, message=FALSE}
# destination airports at altitude of more than 6000 feet
dest_air <- filter(airports, alt > 6000)

flights_dest <- semi_join( flights, dest_air, by= c( "dest" = "faa")) %>%
              group_by(dest)%>%
              count()

dest_airports <- left_join(flights_dest, airports, by= c("dest" = "faa")) %>%
                select(dest, n, name)%>%
                arrange(desc(n))

kbl <- knitr::kable(dest_airports, col.names = c("Airport code", "Number of Flights","Airport name"),format = "html" )
kableExtra::kable_styling(kbl, full_width = FALSE)
```

## 7. Cancellations and Destination Location

```{r, echo=FALSE, message=FALSE, warning=FALSE}
# first 3 months: number of flights, avg arrival delay, prop of cancelled flights to each destination
f3m <- flights %>%
  filter(month %in% c(1,2,3))%>%
  group_by(dest)%>%
  summarize(
    num_flight = n(),
    avg_arr_delay = mean(arr_delay, na.rm = TRUE),
    prop_cancel = mean(is.na(arr_time) & is.na(dep_time)))%>%
  arrange(desc(num_flight)) %>%
  slice_max(num_flight, n = 50)

ap <- inner_join( f3m,airports, by = c("dest"= "faa"))

ggplot(ap)+
  geom_point(aes(x = lon, y = lat, size = prop_cancel)) +
    borders("state") +
    coord_map() +
    labs( size = "Proportion of Canceled Flights",
          title = "Top 50 destinations and the proportion of canceled flights")+
    theme_void()
```

- Airports in the eastern part of the country show higher proportions of canceled flights than the western part.

```{r, echo=FALSE, message=FALSE , warning=FALSE}
ap$delay <- ifelse (ap$avg_arr_delay > 20, "Delayed", "Not Delayed")

ggplot(ap)+
  geom_point(aes(x = lon, y = lat, size = prop_cancel, color = delay)) +
    borders("state") +
    coord_map() +
    labs( size = "Proportion of Canceled Flights", color = "Delay",
          title = "The average arrival delay encoded using color")+
    theme_void()

ggplot(ap)+
  geom_point(aes(x = lon, y = lat, size = prop_cancel, shape = delay)) +
    borders("state") +
    coord_map() +
    labs( size = "Proportion of Canceled Flights", shape = "Delay",
          title = "The average arrival delay encoded using shape")+
    theme_void()
```

- There are 2 airports where the average arrival delay is more than 20 minutes; they're easier to spot when encoded with color.
- Comparing color and shape for encoding whether the average arrival delay is greater than 20 minutes:
    - With color, it's easier to see the difference between delayed and least-delayed destinations, though with more than 2 levels it could become harder to differentiate between colors.
    - With shape, it's easy to tell differences at a glance, but with more levels it can still be hard to differentiate.

## 8. Arrival vs. Departure Delay Relationship

```{r, echo=FALSE, message=FALSE, warning=FALSE }
nyc <- sample_frac(flights, 0.10)
ggplot(nyc, aes( x = dep_delay, y = arr_delay)) +
    geom_point(alpha = 0.1, color = "blue", size = 0.1) +
    geom_smooth(method = "lm", se = FALSE, color = "red") +
    labs(title = "Scatterplot of arrival delay against departure delay ",
       x = "Departure delay", y = "Arrival delay")
```

- The plot shows that the majority of flights have a departure delay between -20 and 50, and there's a strong positive linear relationship between departure delay and arrival delay.

```{r, echo=FALSE, message=FALSE, warning=FALSE }
fl <- filter(nyc, dep_delay <= 30)
ggplot(fl, aes( x = dep_delay, y = arr_delay)) +
    geom_point(alpha = 0.1, color = "blue", size = 0.2, position = "jitter") +
    geom_smooth(method = "lm", se = FALSE, color = "red") +
    labs(title = "Scatterplot of arrival delay against departure delay of at most 30 minutes ",
       x = "Departure delay", y = "Arrival delay")
```

- Focusing on flights with a departure delay of at most 30 minutes:
    - The majority of flights have a departure delay between -10 and 0, whereas for arrival delay the majority falls between -50 and 25.
    - There's a weak positive linear relationship between departure delay and arrival delay.

```{r, echo=FALSE, message=FALSE, warning=FALSE }
p <- ggplot(fl, aes( x = dep_delay, y = arr_delay)) +
    geom_point(alpha = 0.1, color = "blue", size = 0.1, position = "jitter")+
    geom_smooth(method = "lm", se = FALSE, color = "yellow") +
    labs(title = "Previous Scatterplot with marginal density plots and 2D density contours", x = "Departure delay", y = "Arrival delay")+
    geom_density_2d(bins = 3, color = "red")

ggMarginal(p, type = "histogram", bins = 50, fill = "lightgrey")
```

- Marginal plots and contours highlight the majority of the dataset, showing clearly that most flights have a departure delay between -10 and 0.

## 9. Wind Speed, Time of Day, and Departure Delays

```{r, echo=FALSE, message=FALSE, warning=FALSE }
weather_flights <- inner_join(weather, flights, by = c("year", "month", "day", "hour", "origin"))

# select flights in June, July, August; remove rows where wind speed or
# departure delay are missing, and rows where wind speed is >= 30 mph
summer_fl <- filter(weather_flights, month %in% c(6,7,8)) %>%
              drop_na(c(wind_speed, dep_delay)) %>%
              filter(wind_speed < 30) %>%
              group_by(wind_speed) %>%
              summarise(
                n = n(),
                prop_delay = mean(dep_delay > 0)
              )

ggplot(summer_fl, aes(x = wind_speed, y = prop_delay, size = n)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE, linetype = "dashed",aes(weight = n)) +
  labs(title = "Scatterplot of proportion delayed against wind speed < 30 mph",
       x = "Wind Speed (mph)", y = "Proportion Delayed", size = "Number of flights") +
  scale_x_continuous(limits = c(0, 30)) +
  theme_classic()

time_fl <- filter(weather_flights, month %in% c(6,7,8)) %>%
              drop_na(c(wind_speed, dep_delay)) %>%
              filter(wind_speed < 30) %>%
              group_by(hour, wind_speed) %>%
              summarise(
                n = n(),
                prop_delay = mean(dep_delay > 0)
              )
```

- There is a strong positive linear relationship between the proportion delayed and wind speed.
- As wind speed exceeds 20 mph, the proportion of delayed flights increases while the number of flights decreases.
- The highest number of flights occurs when wind speed is less than 15 mph.

```{r, echo=FALSE, message=FALSE, warning=FALSE }
ggplot(time_fl, aes(x = hour, y = prop_delay)) +
  geom_point(aes(size = n)) +
  geom_smooth(method = "lm", se = FALSE, aes(weight = n)) +
  labs(title = "Faceted coplot of proportion delayed against hour conditioned on wind speed",
       x = "Hour", y = "Proportion Delayed", size = "Number of flights") +
  facet_wrap(~ cut_number(wind_speed, 6)) +
  theme_classic()
```

- There is a strong positive linear relationship between proportion delayed and hour when conditioned on wind speed.
- Few flights occur after 8pm, and some of those show a higher proportion of delays.
- There are few flights in the highest wind speed range (19.6–27.6 mph) at any hour.
- There is a higher number of flights when wind speed is between 5.75 and 12.7 mph.
