---
output:
  github_document:
    keep_html: true
    pandoc_args: ["--self-contained", "--mathjax"]
always_allow_html: true
---

**Jocelyne Horanituze**

Sep 10, 2026

This section looks at two unrelated datasets through the lens of distribution and diagnostic plots: self-reported height data (density and eCDF plots, illustrating rounding/heaping bias), and rubber abrasion-loss testing data (scatterplot matrix and faceted coplots examining hardness, tensile strength, and abrasion loss).

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(ggplot2)
library(dplyr)
library(GGally)
```

## 1. Self-Reported Heights

```{r, echo=FALSE, message=FALSE}
data(heights, package = "dslabs")

ggplot(heights, aes(x = height, fill = sex)) +
  geom_density(alpha = 0.5) +
  labs(x = "Height", fill = "Sex", title = "Density Plot of Heights for Males and Females") +
  theme_get()

ggplot(heights)+
  stat_ecdf(aes(x = height, color = sex))+
  labs(y = "Cumulative Frequency", title = "eCDF Plot of Heights for Males and Females")+
  theme_get()
```

- With the density plot:
    - We can easily see the overall distribution of both groups.
    - The small number in the female group has the highest height of around 64 inches, while for the male group the frequency is higher at the highest height compared to the female group.
- With the eCDF plot:
    - It's easier to see the cumulative frequency for each group.
    - The highest number in the female group has a height less than 70 inches, while in the male group a small number have a height below 65 inches.

## 2. Abrasion Loss in Rubber Samples

```{r, echo=FALSE, message=FALSE }
abrasion <- read.csv("abrasion.csv")

ggpairs(abrasion,
        lower = list(continuous = wrap("points",size = 1))) +
        labs(title = "A scatterplot matrix of Abrasion data")
```

- There is a strong negative linear relationship between hardness and abrasion.loss (correlation of -0.738): as hardness increases, abrasion.loss decreases.
- There is a weak negative linear relationship between tensile.strength and abrasion.loss (correlation of -0.299).

```{r, echo=FALSE, message=FALSE, warning=FALSE}
ggplot(abrasion, aes( x = tensile.strength, y = abrasion.loss)) +
    geom_point(data = mutate(abrasion , C = NULL),color = "gray", size = 0.5) +
    geom_point(alpha = 0.5, color = "blue", size = 1) +
    facet_wrap(~ cut_number(hardness, 4)) +
    geom_smooth(method = "lm", se = FALSE, color = "red") +
    labs(title = "Faceted Coplot of Abrasion Loss Against Tensile Strength conditioned on 4 levels of hardness",
       x = "Tensile Strength", y = "Abrasion Loss")
```

- Regardless of the level of hardness, the relationship between tensile.strength and abrasion.loss is negative, meaning that as tensile strength increases, abrasion loss decreases.
    - The relationship is strong and negative at lower levels of hardness, while at higher levels of hardness it is negative but weaker.
