---
author: "Jocelyne Horanituze"
date: "`r format(Sys.Date(), '%B %e, %Y')`"
output:
  github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


In this file, we will be using the *faithful* data set in R.

## 1. Average Waiting Time Between Eruptions

```{r}
with(faithful, mean(waiting))
``` 

- The average waiting time between eruptions of the *Old Faithful* geyser in the data set `faithful` is `r with(faithful, mean(waiting))` minutes.


## 2. First Four Eruption Durations

```{r}
with(faithful, mean(eruptions[1:4]))
``` 

- The average duration of the first 4 eruptions of the *Old Faithful* geyser in the data set `faithful` is `r with(faithful, mean(eruptions[1:4]))` minutes.

## 3. First Five Records of the Eruptions Data

- The first five rows of the `faithful` data frame.

```{r, echo = FALSE}
d <-data.frame(head(faithful, n=5))
kbl <- knitr::kable(d, format = "html")
kableExtra::kable_styling(kbl, full_width = FALSE)
```

## 4. Histogram of Eruption Durations

```{r}
with(faithful, hist(eruptions, labels= TRUE, main = "Histogram of Eruption Durations",
      xlab = "Eruption Durations"))
```

- More than 100 eruptions lasted at least 4 minutes.
- Our data shows some variability; some eruption durations are shorter (less than 3 minutes)
  while others are long. 

