---
author: "Jocelyne Horanituze /"
date: "June 15, 2023"
output:
  github_document:
    keep_html: true
    pandoc_args: ["--self-contained", "--mathjax"]
always_allow_html: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

In this file, we will be using 

- the *Gapminder Data* data from dslabs package in R.
        - It's a health and income outcomes for 184 countries from 1960 - 2016. 
- The NOAA Earth System Research Laboratory [provides data](https://gml.noaa.gov/ccgg/trends/) on monthly mean carbon dioxide (CO2) measured at the Mauna Loa Observatory in Hawaii.

## 1. Average Infant Mortality Rate by Continent in 2015

```{r, echo = FALSE, message = FALSE, warning=FALSE}
library(dplyr)
data(gapminder, package = "dslabs")
d <-data.frame( gapminder %>%
                filter(year ==2015)%>%
                group_by(continent)%>%
                summarise(Average_infant_mortality_rate = mean(infant_mortality, na.rm = TRUE))%>%
                arrange(desc(Average_infant_mortality_rate)))
kbl <- knitr::kable(d, format = "html")
kableExtra::kable_styling(kbl, full_width = FALSE)
```

- Average infant mortality rate, in deaths per 1000, for each continent for the year 2015.
  - Africa shows the highest infant mortality rate of 47.9 , and Europe has lowest infant 
  mortality rate of 4.5.

## 2. Average Infant Mortality Rate by Continent Over the Years

```{r, echo = FALSE, message=FALSE, warning=FALSE}
library(dplyr)
library(ggplot2)
d <-data.frame( gapminder %>%
                group_by(continent, year)%>%
                summarise(Average_infant_mortality_rate = mean(infant_mortality, na.rm = TRUE)))
ggplot(d) + 
     ggtitle("Average Infant Mortality rate by Continent Over the Years")+
     theme(plot.title = element_text(hjust = 0.5))+
     geom_line(aes( x = year,
                   y= Average_infant_mortality_rate,
                   group = (continent),
                   color = continent)) +
    ylab("Average Infant Mortality Rate")
              
      
```

- Average infant mortality rate, in deaths per 1000, for each continent over the years.

- The graph shows an overall decrease in infant mortality rate over the years across continents. 
  - Europe showed the lowest infant mortality rate among all continents over the years, and Africa had the highest infant mortality rate.

## 3. Mauna Loa Atmospheric CO2 Concentration

```{r, echo = FALSE , message = FALSE, warning= FALSE}
library(tidyr)
co2_2022 <- read.csv("co2-2022.csv", na.strings = "***")
lco2_2022<- pivot_longer(co2_2022,
                         -year,
                         names_to = "month",
                         values_to = "co2")
Avg_lco2_2022 <- data.frame(lco2_2022 %>% 
                            group_by(year) %>% 
                            summarise(Yearly_average_co2 = mean(co2, na.rm = TRUE)))

ggplot(Avg_lco2_2022) + 
     ggtitle("Yearly Average Mauna Loa CO2 Concentration Over the Years In Parts Per Million")+
     theme(plot.title = element_text(hjust = 0.5))+
     geom_line(aes( x = year,
                   y= Yearly_average_co2)) +
   ylab("co2")
    
```

- The graph shows an overall increase in Atmospheric CO2 Concentration over the years.
- The CO2 concentration increase from 1960 - 2000, is almost the same as the past 20 years:
    - CO2 Concentration has increased faster over the past 20 years than it did, over 40 years
      from 1960 - 2000.














